---
title: Riddler Puzzle 1
author: Adam Sampson
date: '2018-08-01'
slug: riddler-solution-1
categories: []
tags:
  - R
  - recursion
  - puzzle
draft: false
---

# Solving The Riddler doodle puzzle

FiveThirtyEight has a set of puzzles they post weekly and this week I came across one I thought would be good for practicing recursive programming. The problem is to figure out what moves will solve a puzzle. 

https://fivethirtyeight.com/features/the-perfect-doodle-puzzle-to-keep-you-busy-during-boring-meetings/

> Riddler Classic
> From Renaud Dubedout, a puzzle perfect for doodling during a boring class or meeting, not that I would ever endorse > that sort of thing:
> 
> Start with an empty 5-by-5 grid of squares, and choose any square you want as your starting square. The rules for > moving through the grid from there are strict:
> 
> You may move exactly three cells horizontally or vertically, or you may move exactly two cells diagonally.
> You are not allowed to visit any cell you already visited.
> You are not allowed to step outside the grid.
> You win if you are able to visit all 25 cells.
> 
> Is it possible to win? If so, how? If not, what are the largest and smallest numbers of squares you can legally visit?

# Solving the problem recursively

This problem is somewhat easy to solve using a recursive function. There are much easier recursive problems to start on, but this one is definitely possible. 

In order to solve this recursively I've run the order of operations backwards from what might seem normal. The function starts by checking whether something is a valid move and by making the move if it is possible. Then it checks to see if the move finishes the puzzle. Then the function creates a list of all possible moves and re-runs itself for each possible move. 

Many approaches that aren't recrusive would be to come up with a list of all possible moves, check the moves, and then repeat. But in recursive programming it is easiest if the last action you perform is to recursively call the function again. This allows the recursive nature of the function to do the solving instead of large logic trees.

## Setting up the code

To set up the code we need an empty starting matrix and a function to try out all of the potential possible moves. After a single move you have multiple new possible moves, branching out at every opportunity. When a move isn't possible that branch dies. This is a perfect example of a recursive problem up to 25 levels deep. 

```{r}
#####
# https://fivethirtyeight.com/features/the-perfect-doodle-puzzle-to-keep-you-busy-during-boring-meetings/
# solving with R.
#---

init_matrix <- matrix(integer(),nrow = 5,ncol = 5)
pos <- list(col=2,row=1)

wander_path <- function(cur.pos, curr.matrix, cur.iter = 1){
  # If the position is not on the board return NULL
  if((cur.pos$col<1) | 
     (cur.pos$col>ncol(curr.matrix)) |
     (cur.pos$row<1) |
     (cur.pos$row>nrow(curr.matrix))
  ) {
    return(NULL)
  }
  
  # If the position is already taken return NULL
  if(!is.na(curr.matrix[cur.pos$row,cur.pos$col])){
    return(NULL)
  }
  
  # It must be a valid move if the code gets here...
  
  # Add the current current position to the matrix of taken positions
  curr.matrix[cur.pos$row,cur.pos$col] <- cur.iter
  
  # If this move fills up the board then return the results of the board
  if(!any(is.na(curr.matrix))){
    return(list(curr.matrix))
  }
  
  # Otherwise create a list of new positions to check
  new.pos <- list(
    # For each horizontal/vertical move
    list(col = cur.pos$col + 3, row = cur.pos$row),
    list(col = cur.pos$col - 3, row = cur.pos$row),
    list(col = cur.pos$col , row = cur.pos$row + 3),
    list(col = cur.pos$col , row = cur.pos$row - 3),
    # For each diagonal move
    list(col = cur.pos$col + 2, row = cur.pos$row + 2),
    list(col = cur.pos$col + 2, row = cur.pos$row - 2),
    list(col = cur.pos$col - 2, row = cur.pos$row + 2),
    list(col = cur.pos$col - 2, row = cur.pos$row - 2)
  )
  
  # Recursively run all the possible next moves, then combine the results
  # into a single list while dropping any NULL results
  out <- do.call(c,lapply(new.pos, wander_path, curr.matrix, cur.iter + 1))
  return(out)
}

test_run <- wander_path(cur.pos = pos, curr.matrix = init_matrix)

# Show the first result
(test_run[[1]])
#     [,1] [,2] [,3] [,4] [,5]
#[1,]   16    1   23   15    2
#[2,]    6   20   12    5   21
#[3,]   24    9   17   25   10
#[4,]   13    4   22   14    3
#[5,]    7   19   11    8   18
```

## Running the function for all possible starting positions

Once a function is created to run all possible moves for a starting position, it is pretty trivial to run the function for all starting positions.

This code takes about a minute and a half to run. If you want to save time you could easily notice that there are only really only 6 unique starting positions (out of the 25). The rest of the starting positions can be calculated by a combination of rotation/mirroring of the 6 unique starting positions. A little bit more on this later.

```{r}
winning_runs <- list()
for(i in 1:5){
  for(j in 1:5){
    winning_runs[paste0(i,",",j)] <- list(wander_path(cur.pos = list(row=i,col=j), curr.matrix = init_matrix))
  }
}

# Let's see the first set of winning moves found
(winning_runs$`1,1`[[1]])
#     [,1] [,2] [,3] [,4] [,5]
#[1,]    1   16   19    2   15
#[2,]   11   22    5   12   21
#[3,]   18    8   25   17    7
#[4,]    4   13   20    3   14
#[5,]   10   23    6    9   24
```

## How many possible solutions for each starting position?

Using the above we can pretty quickly calculate the number of solutions for each starting position by adding up the lengths of each list. This can be stored in a matrix. 

```{r}
# Let's see how many winning paths there are for each starting position
win_counts <- matrix(integer(),nrow = 5,ncol = 5)
for(i in 1:5){
  for(j in 1:5){
    win_counts[i,j] <- length(winning_runs[[paste0(i,",",j)]])
  }
}

# Let's see how many winning paths are available for each starting position
(win_counts)
#     [,1] [,2] [,3] [,4] [,5]
#[1,]  552  548  552  548  552
#[2,]  548  412  400  412  548
#[3,]  552  400  352  400  552
#[4,]  548  412  400  412  548
#[5,]  552  548  552  548  552
```

```{r}
# Many of these are mirrored, let's only look at the starting positions that can't
# be created by rotating or mirroring the board.
#       [,1] [,2] [,3] [,4] [,5]
# [1,]  552  548  552
# [2,]       412  400
# [3,]            352
# [4,]  
# [5,]
# But note that within each starting point many of the paths taken are mirrors of other paths...
```


# Conclusion

R is easily able to handle this problem recursively. It takes a little over a minute to run because you can get very far down the tree (and all the way to the end) on many of the branches. 