---
title: "Riddler Solution 1"
author: "Adam Sampson"
output: html_document
---

# Solving The Riddler doodle puzzle

FiveThirtyEight has a set of puzzles they post weekly and this week I came across one I thought would be good for practicing recursive programming. The problem is to figure out what moves will solve a puzzle. 

> Riddler Classic
> From Renaud Dubedout, a puzzle perfect for doodling during a boring class or meeting, not that I would ever endorse > that sort of thing:
> 
> Start with an empty 5-by-5 grid of squares, and choose any square you want as your starting square. The rules for > moving through the grid from there are strict:
> 
> You may move exactly three cells horizontally or vertically, or you may move exactly two cells diagonally.
> You are not allowed to visit any cell you already visited.
> You are not allowed to step outside the grid.
> You win if you are able to visit all 25 cells.
> 
> Is it possible to win? If so, how? If not, what are the largest and smallest numbers of squares you can legally visit?

# Interpreting the rules

There are two ways to interpret the English used to describe the rules. This has to do with how you interpret the comma and the or clauses. If you interpret the comma as seperating two clauses then you get a different interpretation than if you interpret the two 'or' clauses as separating 3 options.

I believe that case one described below is the correct interpretation for how the game should be played, but since we are debating the semantics of the English language I would suggest we re-write the question for clarity and I'll solve both. 

1. On any given move you may move horizontally 3, vertically 3, or diagonally 2
2. For the entire game you may either move horizontally or vertically.

     a. On any move you may move either horizontally or diagonally
     
     b. On any move you may move either vertically or diagonally

# Interpretation 1 of the rules...

1. On any given move you may move horizontally 3, vertically 3, or diagonally 2

## Setting up the code

To set up the code we need an empty starting matrix and a function to try out all of the potential possible moves. After a single move you have multiple new possible moves, branching out at every opportunity. When a move isn't possible that branch dies. This is a perfect example of a recursive problem up to 25 levels deep. 

```{r setup, include=FALSE}
#####
# https://fivethirtyeight.com/features/the-perfect-doodle-puzzle-to-keep-you-busy-during-boring-meetings/
# solving with R.
#---

init_matrix <- matrix(integer(),nrow = 5,ncol = 5)
pos <- list(col=2,row=1)

wander_path <- function(cur.pos, curr.matrix, cur.iter = 1){
  # If the position is not on the board return NULL
  if((cur.pos$col<1) | 
     (cur.pos$col>ncol(curr.matrix)) |
     (cur.pos$row<1) |
     (cur.pos$row>nrow(curr.matrix))
  ) {
    return(NULL)
  }
  
  # If the position is already taken return NULL
  if(!is.na(curr.matrix[cur.pos$row,cur.pos$col])){
    return(NULL)
  }
  
  # It must be a valid move if the code gets here...
  
  # Add the current current position to the matrix of taken positions
  curr.matrix[cur.pos$row,cur.pos$col] <- cur.iter
  
  # If this move fills up the board then return the results of the board
  if(!any(is.na(curr.matrix))){
    return(list(curr.matrix))
  }
  
  # Otherwise create a list of new positions to check
  new.pos <- list(
    # For each horizontal/vertical move
    list(col = cur.pos$col + 3, row = cur.pos$row),
    list(col = cur.pos$col - 3, row = cur.pos$row),
    list(col = cur.pos$col , row = cur.pos$row + 3),
    list(col = cur.pos$col , row = cur.pos$row - 3),
    # For each diagonal move
    list(col = cur.pos$col + 2, row = cur.pos$row + 2),
    list(col = cur.pos$col + 2, row = cur.pos$row - 2),
    list(col = cur.pos$col - 2, row = cur.pos$row + 2),
    list(col = cur.pos$col - 2, row = cur.pos$row - 2)
  )
  
  # Recursively run all the possible next moves, then combine the results
  # into a single list while dropping any NULL results
  out <- do.call(c,lapply(new.pos, wander_path, curr.matrix, cur.iter + 1))
  return(out)
}

test_run <- wander_path(cur.pos = pos, curr.matrix = init_matrix)

# Show the first result
(test_run[[1]])
#     [,1] [,2] [,3] [,4] [,5]
#[1,]   16    1   23   15    2
#[2,]    6   20   12    5   21
#[3,]   24    9   17   25   10
#[4,]   13    4   22   14    3
#[5,]    7   19   11    8   18
```

## Running the function for all possible starting positions

Once a function is created to run all possible moves for a starting position, it is pretty trivial to run the function for all starting positions.

This code takes about a minute and a half to run. If you want to save time you could easily notice that there are only really only 6 unique starting positions (out of the 25). The rest of the starting positions can be calculated by a combination of rotation/mirroring of the 6 unique starting positions. A little bit more on this later.

```{r setup, include=FALSE}
winning_runs <- list()
for(i in 1:5){
  for(j in 1:5){
    winning_runs[paste0(i,",",j)] <- list(wander_path(cur.pos = list(row=i,col=j), curr.matrix = init_matrix))
  }
}

# Let's see the first set of winning moves found
(winning_runs$`1,1`[[1]])
#     [,1] [,2] [,3] [,4] [,5]
#[1,]    1   16   19    2   15
#[2,]   11   22    5   12   21
#[3,]   18    8   25   17    7
#[4,]    4   13   20    3   14
#[5,]   10   23    6    9   24
```

## How many possible solutions for each starting position?

Using the above we can pretty quickly calculate the number of solutions for each starting position by adding up the lengths of each list. This can be stored in a matrix. 

```{r setup, include=FALSE}
# Let's see how many winning paths there are for each starting position
win_counts <- matrix(integer(),nrow = 5,ncol = 5)
for(i in 1:5){
  for(j in 1:5){
    win_counts[i,j] <- length(winning_runs[[paste0(i,",",j)]])
  }
}

# Let's see how many winning paths are available for each starting position
(win_counts)
#     [,1] [,2] [,3] [,4] [,5]
#[1,]  552  548  552  548  552
#[2,]  548  412  400  412  548
#[3,]  552  400  352  400  552
#[4,]  548  412  400  412  548
#[5,]  552  548  552  548  552
```

```{r setup, include=FALSE}
# Many of these are mirrored, let's only look at the starting positions that can't
# be created by rotating or mirroring the board.
#       [,1] [,2] [,3] [,4] [,5]
# [1,]  552  548  552
# [2,]       412  400
# [3,]            352
# [4,]  
# [5,]
# But note that within each starting point many of the paths taken are mirrors of other paths...
```


# Interpretation 2 of the rules...

2. For the entire game you may either move horizontally or vertically.

     a. On any move you may move either horizontally or diagonally
     
     b. On any move you may move either vertically or diagonally
     
Since a and b are simply transpositions of each other, after calculating all staring positions for one we will have the translated results for the other. In other words...we only have to calculate one of the two. 

## Setting up the code

I've made some modifications to the code for this round. When a branch dies, instead of returning NULL we return what the depth of the branch was where it died. 

```{r}
# Assuming the rules meant you have to choose between horizontal or vertical
# for the entire game...choosing horizontal or diagonal for each move
wander_path2 <- function(cur.pos, curr.matrix, cur.iter = 1){
  # If the position is not on the board return NULL
  if((cur.pos$col<1) | 
     (cur.pos$col>ncol(curr.matrix)) |
     (cur.pos$row<1) |
     (cur.pos$row>nrow(curr.matrix))
  ) {
    return(NULL)
  }
  
  # If the position is already taken return how far the branch made it
  if(!is.na(curr.matrix[cur.pos$row,cur.pos$col])){
    return(cur.iter - 1)
  }
  
  # It must be a valid move if the code gets here...
  
  # Add the current current position to the matrix of taken positions
  curr.matrix[cur.pos$row,cur.pos$col] <- cur.iter
  
  # If this move fills up the board then return the results of the board
  if(!any(is.na(curr.matrix))){
    return(list(curr.matrix))
  }
  
  # Otherwise create a list of new positions to check
  new.pos <- list(
    # For each horizontal/vertical move
    list(col = cur.pos$col + 3, row = cur.pos$row),
    list(col = cur.pos$col - 3, row = cur.pos$row),
    # For each diagonal move
    list(col = cur.pos$col + 2, row = cur.pos$row + 2),
    list(col = cur.pos$col + 2, row = cur.pos$row - 2),
    list(col = cur.pos$col - 2, row = cur.pos$row + 2),
    list(col = cur.pos$col - 2, row = cur.pos$row - 2)
  )
  
  # Recursively run all the possible next moves, then combine the results
  # into a single list while dropping any NULL results
  out <- do.call(c,lapply(new.pos, wander_path2, curr.matrix, cur.iter + 1))
  return(out)
}

test_run2 <- wander_path2(cur.pos = pos, curr.matrix = init_matrix)
max(test_run2)
# [1] 12
```

## Running the function for all possible starting positions

There is a slight modification to this code to output the farthest number. If there are any winning moves the function will output list(s) with the winning solution and another vector with the farthest reached by non-winning branches.

```{r}
winning_runs2 <- list()
for(i in 1:5){
  for(j in 1:5){
    winning_runs2[paste0(i,",",j)] <- list(wander_path2(cur.pos = list(row=i,col=j), curr.matrix = init_matrix))
  }
}

# If there are any moves the function will return a $ 5,5:List of ... containing
# a list of at least one matrix num [1:5, 1:5]. Otherwise it will just contain 25 
# containing a single vector of numbers. 
str(winning_runs2)
# List of 25
# $ 1,1: num [1:174] 2 4 5 7 9 10 12 12 11 12 ...
# $ 1,2: num [1:174] 2 5 7 10 12 12 11 12 12 11 ...
# $ 1,3: num [1:196] 3 5 8 10 11 11 10 9 11 11 ...
# $ 1,4: num [1:174] 2 5 7 10 12 12 11 10 9 10 ...
# $ 1,5: num [1:174] 2 4 7 9 12 12 11 10 10 12 ...
# $ 2,1: num [1:20] 2 4 5 7 9 10 10 8 6 3 ...
# $ 2,2: num [1:20] 2 3 5 7 8 10 10 9 6 4 ...
# $ 2,3: num [1:20] 3 5 8 10 10 9 7 6 4 2 ...
# $ 2,4: num [1:20] 2 5 7 10 10 9 8 6 4 3 ...
# $ 2,5: num [1:20] 2 4 7 9 10 10 8 6 5 3 ...
# $ 3,1: num [1:147] 2 4 5 7 7 6 7 9 10 10 ...
# $ 3,2: num [1:139] 2 3 5 7 8 10 10 9 10 10 ...
# $ 3,3: num [1:116] 3 5 8 10 10 9 10 10 9 7 ...
# $ 3,4: num [1:139] 2 5 7 10 10 9 10 10 9 8 ...
# $ 3,5: num [1:147] 2 4 7 9 10 10 10 10 8 10 ...
# $ 4,1: num [1:20] 2 4 5 7 9 10 10 8 6 3 ...
# $ 4,2: num [1:20] 2 3 5 7 8 10 10 9 6 4 ...
# $ 4,3: num [1:20] 3 5 8 10 10 9 7 6 4 2 ...
# $ 4,4: num [1:20] 2 5 7 10 10 9 8 6 4 3 ...
# $ 4,5: num [1:20] 2 4 7 9 10 10 8 6 5 3 ...
# $ 5,1: num [1:174] 2 4 5 7 9 10 12 12 11 10 ...
# $ 5,2: num [1:174] 2 3 5 7 10 12 12 11 12 12 ...
# $ 5,3: num [1:196] 3 5 8 10 10 11 11 9 11 11 ...
# $ 5,4: num [1:174] 2 5 7 10 10 12 12 11 9 10 ...
# $ 5,5: num [1:174] 2 4 7 9 10 12 12 11 10 12 ...

# Let's see the the farthest ever reached
max(unlist(lapply(winning_runs2, max)))
# [1] 13
```

## How far can you get from each possible starting position?

```{r}
# Let's see how far you can get for each starting position
max_move_counts2 <- matrix(integer(),nrow = 5,ncol = 5)
for(i in 1:5){
  for(j in 1:5){
    max_move_counts2[i,j] <- max(winning_runs2[[paste0(i,",",j)]])
  }
}

max_move_counts2
#       [,1] [,2] [,3] [,4] [,5]
# [1,]   13   12   13   12   13
# [2,]   10   10   10   10   10
# [3,]   12   11   10   11   12
# [4,]   10   10   10   10   10
# [5,]   13   12   13   12   13
```

Depending on the starting position you choose you can make it from 10 to 13 moves before failing using interpretation 2. 

# Conclusion

R is easily able to handle this problem recursively. Method 1 takes a little over a minute to run because you can get very far down the tree (and all the way to the end) on many of the branches. Method 2 is exceptionally fast because it is not possible to get very far down the tree. 

Also, reiterating, a better problem description would be nice.
